The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-04-15 src/ms.tex
fix capitalization in project vs Project

\begin{abstract}
The \astropy Project supports and fosters the development of open-source and openly developed
\python packages that provide commonly needed functionality to the astronomical
community.
A key element of the \astropy Project is the core package \astropypkg, which serves as the
foundation for more specialized projects and packages.
In this article, we summarize key features in the core package as of the recent major
release, version 5.0, and provide major updates for the project.
We then discuss supporting a broader ecosystem of interoperable packages,
including connections with several astronomical observatories and missions.
We also revisit the future outlook of the \astropy Project and the current
status of Learn Astropy.
We conclude by raising and discussing the current and future challenges facing the project.
\end{abstract}

@@

The FITS standard was extended to rigorously describe time coordinates in the
WCS framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{ To learn more about this project, please
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.} the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a mixin column. For
backward compatibility, this feature may be turned off.

@@

Since its inception, the most active contributors to the \astropypkg package
have spanned a range of career stages and paths.
For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
career stage/path at the midpoint of the three time eras defined in the previous
paragraph for all contributors with more than 50 commits to the \astropypkg
package within each era.
Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
contributors moved between these categories (i.e., have graduated as students to
become research staff/faculty or software engineers), but also a number of
early contributors moved on to other positions and stopped contributing to the
core \astropypkg package.
Given this, it is encouraging that the numbers in each career stage/path have
stayed relatively constant over the history of \astropy (with some showing
growth), as new, active contributors have generally joined and stepped in to
fill top-contributor roles within the project.
On the other hand, the total number of top contributors has not grown
substantially, reflecting one of our principal challenges: sustaining a pipeline
from users to contributors to maintainers of the package.
One other thing to note in this context is that in all eras, 50--60\% of the
software engineers who contributed to \astropypkg were (or are) employed by
STScI, whose leadership has consistently provided resources (financial and
personnel) towards sustaining \astropy.

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
        \caption{
            \textbf{Top panel:} The number of Git commits per month (defined as
            consecutive 30 day windows) merged into the \astropypkg core
            package over the whole history of the project (up to end of 2021,
            v5.0).
            The dark line shows the mean rate using a 1 year rolling window, but
            the under-plotted gray line shows the raw monthly rate.
            The raw number clearly peaks just before feature freezes (vertical
            lines) as a flurry of activity leading up to each release, but in
            recent years the mean rate has remained steady at about $\sim$300
            commits per month.
            \textbf{Bottom panel:} Similar to above, but now showing the number
            of unique Git committers who contribute to the \astropypkg core
            package each month (again defined as consecutive 30 day windows).
            While this also shows some variation depending on proximity to
            feature freezes, in recent years the mean rate of unique
            contributors has remained consistent at about $\sim$22 committers
            per month.
        }
        \label{fig:contributor-summary:per-month}
    \end{centering}
\end{figure}

@@

With support from the Moore Foundation, the \astropy Project was
allocated funding to support mentoring programs. In 2020 a call was
made to submit proposals for IDE (Inclusion, Diversity, and
Empowerment) initiatives for project-wide consideration on
\github. This process was deemed the most ``open'' because it allowed
for community-wide feedback to focus and improve proposal
initiatives. Two highlighted programs that were selected and implemented via
this process are described below.

@@

In addition to the supported mentoring initiatives, the \astropy
project as a whole has taken steps to examine representation of
marginalized groups within the project and search for avenues of
improvement.

@@

\begin{description}
    \item[Moore] The Gordon and Betty Moore Foundation awarded the \astropy
    Project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project.'' Over a three-year term, this grant supported
    sustenance of the Project by providing funds for (1) the development and
    maintenance of \astropypkg and its infrastructure, provided to existing
    project members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the
    formalization of a governance structure (see
    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
    diversity, and inclusion efforts in the project (see
    Section~\ref{sec:project-ide}). Some of this money supported travel for
    meetings, conferences, and workshops. This generous grant helped reduce
    reliance on volunteer-driven maintenance of the openly-developed package and
    paved the way to acquire funding from US federal agencies, representing a
    major milestone for the success of the project.

    \item[NASA] A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    \astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the project by encouraging further
    development of the package ecosystem and community engagement.
\end{description}

@@

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

Sphinx is a common and useful tool for generating documentation for \python
packages. The Astropy project maintains a default Sphinx configuration along
with Astropy-specific extensions that can be easily added to community
projects via the
\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}}
meta-package. This tool provides a pre-configured Sphinx setup compatible with
Astropy projects, which includes several extensions useful for generating API
documentation
(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
allowing for \package{numpy} docstring parsing
(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
handling
(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
documentation testing support
(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
and providing a custom documentation theme ideal for analysis packages
(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).

@@

Beyond developing and serving educational content, the \astropy Project has been
conducting workshops at winter meetings of the American Astronomical Society
since AAS 225 in January 2015.
Up to the start of the coronavirus pandemic, these were full-day in-person
workshops with as many as 90 participants and a dozen facilitators from the
project.
During the pandemic, these workshops were moved to an online format and split
into basic and advanced sessions.
Additionally, beginning with the AAS 238 online meeting, the workshops have been
expanded to Summer AAS meetings.
The Learn team finds that the workshop audience is best found at AAS meetings as
opposed to more general \python meetings, as the content tends to be more
applicable for students and researchers in astronomy and astrophysics.

@@

The
\href{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}{Astropy Roadmap} is a document in the \texttt{astropy-project} \github
repository within the \astropy organization that captures high-level actionable
items that the Astropy project aims to undertake to improve the health and
stability of the project. It is a static document that is revisited regularly at
the Astropy coordination meetings, intended to keep track of progress and write
new versions as needed. There is a related project board linked to the Roadmap
document that holds specific issues and efforts related to Roadmap items. The
project board is a living document that is continually updated as work is
planned, assigned, and completed.

@@

  The \astropy project aims to be a project that is made up of members at all
  levels (users to contributors to coordinators) with a diverse set of skills,
  backgrounds, and experiences.

@@

  the ability to contribute to the project.

@@

The Astropy Project has continued to serve as an active, community-oriented
software project with a wide reach and growing impact, both within astronomy and
astrophysics and in broader scientific software communities.
The Project is in the midst of setting up infrastructure to ensure its long-term
sustainability and to continue healthy growth of the software ecosystem,
developer, and user communities that interface with the Project.
For example, over the last years, we have formalized and implemented a new
governance model for the Astropy Project (Section~\ref{sec:project-governance}),
have continued to engage a growing pool of active and top contributors
(Section~\ref{sec:project-contributors}), and have begun to receive significant
formal funding for the project itself (Section~\ref{sec:project-funding}).
The Astropy Project is also expanding its educational and community-focused
initiatives by soliciting and funding initiatives in inclusion, diversity, and
equity (Section~\ref{sec:project-ide}) as well as through the Learn Astropy initiative,
which aims to make tutorials and educational content that highlight software
tools for astronomy more discoverable and accessible (Section~\ref{sec:learn}).
We expect to continue in all of these threads over the coming years as we aim to
make the Astropy Project more sustainable and a larger, more welcoming community
for open-source software users and enthusiasts in astronomy.

@@

We would like to thank the members of the community who have contributed to
\astropy, opened issues and provided feedback, and supported the
project in a number of different ways.

>>>>

\begin{abstract}
The \astropy Project supports and fosters the development of open-source and openly developed
\python packages that provide commonly needed functionality to the astronomical
community.
A key element of the \astropy Project is the core package \astropypkg, which serves as the
foundation for more specialized projects and packages.
In this article, we summarize key features in the core package as of the recent major
release, version 5.0, and provide major updates for the Project.
We then discuss supporting a broader ecosystem of interoperable packages,
including connections with several astronomical observatories and missions.
We also revisit the future outlook of the \astropy Project and the current
status of Learn Astropy.
We conclude by raising and discussing the current and future challenges facing the Project.
\end{abstract}

@@

The FITS standard was extended to rigorously describe time coordinates in the
WCS framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{To learn more about this project, please
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.} the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a mixin column. For
backward compatibility, this feature may be turned off.

@@

Since its inception, the most active contributors to the \astropypkg package
have spanned a range of career stages and paths.
For example, Figure~\ref{fig:contributor-summary:top} shows a breakdown of the
career stage/path at the midpoint of the three time eras defined in the previous
paragraph for all contributors with more than 50 commits to the \astropypkg
package within each era.
Over the $\sim$6 years between the ``early'' and ``recent'' eras, many
contributors moved between these categories (i.e., have graduated as students to
become research staff/faculty or software engineers), but also a number of
early contributors moved on to other positions and stopped contributing to the
core \astropypkg package.
Given this, it is encouraging that the numbers in each career stage/path have
stayed relatively constant over the history of \astropy (with some showing
growth), as new, active contributors have generally joined and stepped in to
fill top-contributor roles within the Project.
On the other hand, the total number of top contributors has not grown
substantially, reflecting one of our principal challenges: sustaining a pipeline
from users to contributors to maintainers of the package.
One other thing to note in this context is that in all eras, 50--60\% of the
software engineers who contributed to \astropypkg were (or are) employed by
STScI, whose leadership has consistently provided resources (financial and
personnel) towards sustaining \astropy.

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
        \caption{
            \textbf{Top panel:} The number of Git commits per month (defined as
            consecutive 30 day windows) merged into the \astropypkg core
            package over the whole history of the Project (up to end of 2021,
            v5.0).
            The dark line shows the mean rate using a 1 year rolling window, but
            the under-plotted gray line shows the raw monthly rate.
            The raw number clearly peaks just before feature freezes (vertical
            lines) as a flurry of activity leading up to each release, but in
            recent years the mean rate has remained steady at about $\sim$300
            commits per month.
            \textbf{Bottom panel:} Similar to above, but now showing the number
            of unique Git committers who contribute to the \astropypkg core
            package each month (again defined as consecutive 30 day windows).
            While this also shows some variation depending on proximity to
            feature freezes, in recent years the mean rate of unique
            contributors has remained consistent at about $\sim$22 committers
            per month.
        }
        \label{fig:contributor-summary:per-month}
    \end{centering}
\end{figure}

@@

With support from the Moore Foundation, the \astropy Project was
allocated funding to support mentoring programs. In 2020 a call was
made to submit proposals for Inclusion, Diversity, and
Empowerment (IDE) initiatives for project-wide consideration on
\github. This process was deemed the most ``open'' because it allowed
for community-wide feedback to focus and improve proposal
initiatives. Two highlighted programs that were selected and implemented via
this process are described below.

@@

In addition to the supported mentoring initiatives, the \astropy
Project as a whole has taken steps to examine representation of
marginalized groups within the Project and search for avenues of
improvement.

@@

\begin{description}
    \item[Moore] The Gordon and Betty Moore Foundation awarded the \astropy
    Project a $\sim$\$900k (US) grant in 2019 for the proposal ``Sustaining and
    Growing the Astropy Project.'' Over a three-year term, this grant supported
    sustenance of the Project by providing funds for (1) the development and
    maintenance of \astropypkg and its infrastructure, provided to existing
    project members and targeted hires, (2) the transition of long-term users to
    contributors and maintainers through mentorship efforts, (3) the
    formalization of a governance structure (see
    Section~\ref{sec:project-governance}), and (4) the improvement of equity,
    diversity, and inclusion efforts in the Project (see
    Section~\ref{sec:project-ide}). Some of this money supported travel for
    meetings, conferences, and workshops. This generous grant helped reduce
    reliance on volunteer-driven maintenance of the openly-developed package and
    paved the way to acquire funding from US federal agencies, representing a
    major milestone for the success of the Project.

    \item[NASA] A successful proposal to the NASA ROSES-2020
    call\footnote{http://solicitation.nasaprs.com/ROSES2020}, section E.7
    (Support for Open Source Tools, Frameworks, and Libraries), granted the
    \astropy Project $\sim$\$600k (US) to support its work. Funded work for a
    term of three years includes (1) project-wide infrastructure maintenance and
    improvements, (2) targeted work on areas of the \astropy core and
    coordinated packages, (3) enhancements to the Learn
    Astropy ecosystem and educational materials (see Section~\ref{sec:learn}),
    and (4) support for the \astropy affiliated packages. These goals
    support the sustenance of the Project by encouraging further
    development of the package ecosystem and community engagement.
\end{description}

@@

The Astropy Project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

Sphinx is a common and useful tool for generating documentation for \python
packages. The Astropy Project maintains a default Sphinx configuration along
with Astropy-specific extensions that can be easily added to community
projects via the
\href{https://github.com/astropy/sphinx-astropy}{\texttt{sphinx-astropy}}
meta-package. This tool provides a pre-configured Sphinx setup compatible with
Astropy projects, which includes several extensions useful for generating API
documentation
(\href{https://github.com/astropy/sphinx-automodapi}{\texttt{sphinx-automodapi}}),
allowing for \package{numpy} docstring parsing
(\href{https://github.com/numpy/numpydoc}{\texttt{numpydoc}}), embedded image
handling
(\href{https://github.com/sphinx-gallery/sphinx-gallery}{\texttt{sphinx-gallery}};
\href{https://github.com/python-pillow}{\texttt{pillow}}), advanced
documentation testing support
(\href{https://github.com/astropy/pytest-doctestplus}{\texttt{pytest-doctestplus}}),
and providing a custom documentation theme ideal for analysis packages
(\href{https://github.com/astropy/sphinx-astropy}{\texttt{astropy-sphinx-theme}}).

@@

Beyond developing and serving educational content, the \astropy Project has been
conducting workshops at winter meetings of the American Astronomical Society
since AAS 225 in January 2015.
Up to the start of the coronavirus pandemic, these were full-day in-person
workshops with as many as 90 participants and a dozen facilitators from the
Project.
During the pandemic, these workshops were moved to an online format and split
into basic and advanced sessions.
Additionally, beginning with the AAS 238 online meeting, the workshops have been
expanded to Summer AAS meetings.
The Learn team finds that the workshop audience is best found at AAS meetings as
opposed to more general \python meetings, as the content tends to be more
applicable for students and researchers in astronomy and astrophysics.

@@

The
\href{https://github.com/astropy/astropy-project/blob/main/roadmap/roadmap.md}{Astropy Roadmap} is a document in the \texttt{astropy-project} \github
repository within the \astropy organization that captures high-level actionable
items that the Astropy Project aims to undertake to improve the health and
stability of the Project. It is a static document that is revisited regularly at
the Astropy coordination meetings, intended to keep track of progress and write
new versions as needed. There is a related project board linked to the Roadmap
document that holds specific issues and efforts related to Roadmap items. The
project board is a living document that is continually updated as work is
planned, assigned, and completed.

@@

  The \astropy Project aims to be made up of members at all levels (users to
  contributors to coordinators) with a diverse set of skills, backgrounds, and
  experiences.

@@

  the ability to contribute to the Project.

@@

The Astropy Project has continued to serve as an active, community-oriented
software project with a wide reach and growing impact, both within astronomy and
astrophysics and in broader scientific software communities.
The Project is in the midst of setting up infrastructure to ensure its long-term
sustainability and to continue healthy growth of the software ecosystem,
developer, and user communities that interface with the Project.
For example, over the last years, we have formalized and implemented a new
governance model for the Astropy Project (Section~\ref{sec:project-governance}),
have continued to engage a growing pool of active and top contributors
(Section~\ref{sec:project-contributors}), and have begun to receive significant
formal funding for the Project itself (Section~\ref{sec:project-funding}).
The Astropy Project is also expanding its educational and community-focused
initiatives by soliciting and funding initiatives in inclusion, diversity, and
equity (Section~\ref{sec:project-ide}) as well as through the Learn Astropy initiative,
which aims to make tutorials and educational content that highlight software
tools for astronomy more discoverable and accessible (Section~\ref{sec:learn}).
We expect to continue in all of these threads over the coming years as we aim to
make the Astropy Project more sustainable and a larger, more welcoming community
for open-source software users and enthusiasts in astronomy.

@@

We would like to thank the members of the community who have contributed to
\astropy, opened issues and provided feedback, and supported the
Project in a number of different ways.
